Processing CDC hypertension and CDV Mortality disease data from adults in the USA exploratory analysis script

### 2025-04-19 code/analysis-code/statistical-analysis-v2-slim.qmd
Part 5 Additions

##Setup

@@

#Data fitting/statistical analysis

##Aim 1: Temporal Trends by Age Group

@@

####1.1 Linear Model - Overall Mortality over time

@@

@@

This model provides a first look at overall temporal trends in mortality rates. The flat trend line suggests that overall mortality rates changed minimally from 2000 to 2019, though high variability exists.

#####1.2 GLM: Race and Sex as predictors
GLM with race and sex as additive predictors to assess overall differences in mortality rates
```{r}
# Examining multiple predictors
glm_model <- glm(mortalities ~ Year + race_ethnicity + sex, data = county_race_sex, family = gaussian())
glm_results <- broom::tidy(glm_model)

@@

The GLM model incorporating race/ethnicity and sex reveals substantial disparities in mortality rates between groups. Black populations consistently show the highest predicted rates across the period.

####1.3 GLM: Year by Race interaction model:
Fit GLM including interaction between year and race to examine race-specific trends over time

@@

saveRDS(county_race_sex, file = here::here("results", "output", "county_race_sex_with_preds.rds"))
```
By including interaction terms, we can evaluate whether different racial/ethnic groups experienced distinct temporal trends. Here, trends diverge sharply: Black mortality rates decline, while rates rise for American Indian/Alaska Native populations.

####1.4: GLM: Year by Age group

GLM with interaction between year and age group to examine differential temporal trends
```{r}
# GLM with interaction: does time trend differ by age group?
glm_age_group <- glm(mortalities ~ Year * age_group, 
                     data = processed_data_no_percent,
                     family = gaussian())

@@

@@

```
This model identifies age-specific trends. Mortality among individuals aged 65+ remains substantially higher and shows modest increases, while younger age groups have lower, flatter trends.

####1.5: Slope comparison by Age group

@@

@@

@@

## Aim 2: Spatial Analysis of Hypertension Mortality Rates
## Evaluating spatial distribution and clustering patterns

@@

###2.1: Static maps by Race:

Converting county data to spatial object using longitude/latitude.
Creating static spatial map for all race groups.
```{r}
#Convert dataset to an SF spatial object
county_race_sex_sf <- county_race_sex %>%
  st_as_sf(coords = c("X_long", "Y_lat"), crs = 4326, remove = FALSE)

@@

###2.2 Combined spatial plot

Combining all spatial race-specific maps into a multi-panel figure.
```{r}

@@

```
These plots are best viewed as images outside of R due to size.

Evaluating spatial clustering of Mortality Rates

### 2.3: Global Moran’s I – All Races Combined

@@

Running Global Moran's I test to evaluate overall spatial autocorrelation.
```{r spatial-weights global-morans-i, message=FALSE}
# Define file paths for output objrects
knn_path <- here::here("results", "output", "knn_neighbors.rds") # Larger, reusable object
lw_path <- here::here("results", "output", "spatial_weights_list.rds") # Larger, reusable object
moran_path <- here::here("results", "output", "global_morans_i.rds") # Smaller, result-level output

@@

# Save result
saveRDS(moran_result, file = moran_path)

@@

Global Moran's I: I = 0.0435, p < 0.001, this indicates weak but statistically significant spatial autocorrelation.
There is weak-to-moderate positive spatial autocorrelation in mortality rates across U.S. counties — counties with high (or low) mortality tend to be near others with similarly high (or low) values. Although the magnitude is small (I = 0.0435), the pattern is considered highly statistically significant based on the p-value.

@@

Running Local Moran's I (LISA) to detect local spatial clusters.
```{r lisa-analysis, message=FALSE}

@@

Creating map of statistically significant LISA clusters (p < 0.05).
```{r}

@@

```
Classifying counties based on LISA cluster type using z-scores.

@@

# Define cluster color palette
cluster_colors <- c(
  "High-High" = "red3",
  "Low-Low" = "blue3",
  "High-Low (Outlier)" = "orange",
  "Low-High (Outlier)" = "purple",
  "Not Significant" = "gray85"
)

# Plot cluster types
lisa_cluster_plot <- ggplot(county_lisa_coords) +
  geom_point(aes(x = long, y = lat, color = lisa_cluster_type), alpha = 0.7, size = 0.75) +
  scale_color_manual(values = cluster_colors) +
  labs(
    title = "Local Spatial Clusters of Mortality (LISA)",
    subtitle = "High–High = Hotspot, Low–Low = Coldspot, Others = Outliers",
    color = "Cluster Type"
  ) +
  coord_equal() +
  theme_minimal(base_size = 12)

@@

ggsave(
  here::here("products", "manuscript", "supplement", "lisa_cluster_type_map.png"),
  plot = lisa_cluster_plot,
  width = 10, height = 6
)

```
Local Moran’s I was computed on the full national dataset using 5-nearest-neighbor spatial weights. Each county was represented by a centroid, and clustering patterns reflect spatial autocorrelation of mortality rates across all years and subgroups.

####Focusing at a State level
Due to the size of the dataset it is challenging to work with the full dataset.
For convenience and because the SouthEast is historically known to have a high prevalence of cardiovasccular mortalities we will focus on Georgia for this next step.
Our previous analysis support that there is a general distribution of cases here, even if there are not distinct hotspots.

Analyzing spatial clusters in Georgia using LISA.
```{r}
# Georgia: Create or load spatial weights
ga_knn_path <- here::here("results", "output", "ga_knn_neighbors.rds")
ga_lw_path  <- here::here("results", "output", "ga_spatial_weights.rds")

@@

if (file.exists(ga_knn_path) && file.exists(ga_lw_path)) {
  message("Reusing saved GA spatial weights and neighbors.")
  ga_knn_nb <- readRDS(ga_knn_path)
  ga_lw <- readRDS(ga_lw_path)
} else {
  message("Generating GA spatial weights.")
  ga_coords <- st_coordinates(county_ga_sf)
  ga_knn_nb <- knn2nb(knearneigh(ga_coords, k = 5))
  ga_lw     <- nb2listw(ga_knn_nb, style = "W")
  saveRDS(ga_knn_nb, file = ga_knn_path)
  saveRDS(ga_lw, file = ga_lw_path)
}

@@

cluster_colors <- c(
  "High-High" = "red3",
  "Low-Low" = "blue3",
  "High-Low (Outlier)" = "orange",
  "Low-High (Outlier)" = "purple",
  "Not Significant" = "gray80"
)

ga_lisa_plot <- ggplot(ga_coords_df) +
  geom_point(aes(x = long, y = lat, color = cluster_type), alpha = 0.8, size = 2) +
  scale_color_manual(values = cluster_colors) +
  labs(
    title = "LISA Cluster Types: Georgia",
    subtitle = "High–High = Hotspot; Low–Low = Coldspot; Others = Outliers",
    color = "Cluster Type"
  ) +
  coord_equal() +
  theme_minimal(base_size = 13)

@@

  # Save for reuse
  saveRDS(ms_knn_nb, file = ms_knn_path)
  saveRDS(ms_lw, file = ms_lw_path)
}

@@

county_ms_sf <- county_ms_sf %>%
  mutate(
    local_I = ms_local_moran[, 1],
    local_p = ms_p,
    mortality_z = ms_z,
    lag_z = ms_lag_z,
    cluster_type = case_when(
      local_p >= 0.05 ~ "Not Significant",
      mortality_z > 0 & lag_z > 0 ~ "High-High",
      mortality_z < 0 & lag_z < 0 ~ "Low-Low",
      mortality_z > 0 & lag_z < 0 ~ "High-Low (Outlier)",
      mortality_z < 0 & lag_z > 0 ~ "Low-High (Outlier)",
      TRUE ~ NA_character_
    )
  )

@@

cluster_colors <- c(
  "High-High" = "red3",
  "Low-Low" = "blue3",
  "High-Low (Outlier)" = "orange",
  "Low-High (Outlier)" = "purple",
  "Not Significant" = "gray80"
)

ms_lisa_plot <- ggplot(ms_coords_df) +
  geom_point(aes(x = long, y = lat, color = cluster_type), alpha = 0.8, size = 2.5) +
  scale_color_manual(values = cluster_colors) +
  labs(
    title = "LISA Cluster Types: Mississippi",
    subtitle = "High–High = Hotspot; Low–Low = Coldspot; Others = Outliers",
    color = "Cluster Type"
  ) +
  coord_equal() +
  theme_minimal(base_size = 13)

@@

ggsave(
  filename = here::here("products", "manuscript", "supplement", "ms_lisa_cluster_plot.png"),
  plot = ms_lisa_plot,
  width = 8,
  height = 6
)

@@

# Add LISA values and classification
county_tx_sf <- county_tx_sf %>%
  mutate(
    local_I = tx_local_moran[, 1],
    local_p = tx_p,
    mortality_z = tx_z,
    lag_z = tx_lag_z,
    cluster_type = case_when(
      local_p >= 0.05 ~ "Not Significant",
      mortality_z > 0 & lag_z > 0 ~ "High-High",
      mortality_z < 0 & lag_z < 0 ~ "Low-Low",
      mortality_z > 0 & lag_z < 0 ~ "High-Low (Outlier)",
      mortality_z < 0 & lag_z > 0 ~ "Low-High (Outlier)",
      TRUE ~ NA_character_
    )
  )

@@

# Cluster colors
cluster_colors <- c(
  "High-High" = "red3",
  "Low-Low" = "blue3",
  "High-Low (Outlier)" = "orange",
  "Low-High (Outlier)" = "purple",
  "Not Significant" = "gray80"
)

# Plot
tx_lisa_plot <- ggplot(tx_coords_df) +
  geom_point(aes(x = long, y = lat, color = cluster_type), alpha = 0.8, size = 2) +
  scale_color_manual(values = cluster_colors) +
  labs(
    title = "LISA Cluster Types: Texas",
    subtitle = "High–High = Hotspot; Low–Low = Coldspot; Others = Outliers",
    color = "Cluster Type"
  ) +
  coord_equal() +
  theme_minimal(base_size = 13)

@@

# Save plot
ggsave(
  filename = here::here("products", "manuscript", "supplement", "tx_lisa_cluster_plot.png"),
  plot = tx_lisa_plot,
  width = 9, height = 6
)
```

@@

# Add results to object
county_ky_sf <- county_ky_sf %>%
  mutate(
    local_I = local_moran_ky[, 1],
    local_p = local_moran_ky[, 5],
    mortality_z = scale(mortalities)[, 1],
    lag_z = lag.listw(ky_lw, mortality_z),
    lisa_cluster = case_when(
      local_p > 0.05 ~ "Not Significant",
      mortality_z > 0 & lag_z > 0 ~ "High-High",
      mortality_z < 0 & lag_z < 0 ~ "Low-Low",
      mortality_z > 0 & lag_z < 0 ~ "High-Low (Outlier)",
      mortality_z < 0 & lag_z > 0 ~ "Low-High (Outlier)",
      TRUE ~ NA_character_
    )
  )

@@

```
the LISA plot shows spatial patterns across all available data, not just one year

@@

2.5 Global Moran's I Summary Table

```{r}
# Already saved or placed earlier i
extract_moran_values <- function(obj, geo) {
  tibble(
    Geography = geo,
    `Moran's I` = round(obj$estimate[["Moran I statistic"]], 4),
    Expectation = round(obj$estimate[["Expectation"]], 4),
    Variance = round(obj$estimate[["Variance"]], 6),
    `Z Statistic` = round(obj$statistic, 4),
    `p-value` = signif(obj$p.value, 4))}

@@

# Format with gt
moran_gt_tbl <- moran_summary %>%
  gt() %>%
  tab_header(
    title = "Global Moran's I Statistics by Geography",
    subtitle = "Spatial Autocorrelation of Mortality Rates"
  ) %>%
  fmt_number(columns = 2:5, decimals = 4) %>%  # Keep decimals for other metrics
  fmt_scientific(columns = vars(`p-value`), decimals = 2) %>%  # Use sci notation for p-values
  cols_label(
    `Moran's I` = "Moran's I",
    `Z Statistic` = "Z Statistic",
    `p-value` = "p-value"
  ) %>%
  tab_options(
    table.font.size = px(12),
    heading.title.font.size = px(14),
    heading.subtitle.font.size = px(12),
    column_labels.font.weight = "bold"
  )

@@

A summary table of Global Moran’s I statistics for the full U.S. dataset and selected states (GA, MS, TX, KY) is shown below, indicating statistically significant spatial autocorrelation in all geographies examined (p < 0.001).

Local Indicators of Spatial Association (LISA) – Statistically Significant Clustering of Mortality Rates in Kentucky.
This map highlights counties with statistically significant spatial clustering of hypertension-related cardiovascular disease (CVD) mortality rates, as detected by Local Moran’s I. Each point represents a county centroid. Counties shown in red are spatial clusters where both the county and its neighbors have high mortality values (high-high) or display significant spatial autocorrelation (e.g., high-low or low-high). Spatial relationships were calculated using 5-nearest-neighbor spatial weights based on county centroids. This analysis includes data from all available years and demographic subgroups; therefore, significance reflects elevated spatial correlation within any combination of groupings, not just overall rates.

Local spatial analysis (LISA) identified a small number of statistically significant clusters in Kentucky, including high-mortality counties adjacent to other high-mortality areas. These patterns support the presence of localized mortality hotspots and may inform targeted investigation or intervention efforts.

##Aim 3: Demographic Disparities

@@

@@

@@

###3.2 Demogrpahic disparities by Age group

@@

##Aim 4: Predict County-level Mortality rates

>>>>

# Setup

@@

# Data fitting/statistical analysis

Summary statistics of baseline mortality by race/ethnicity, sex, and age group are provided later (between Aim 3 and 4) to contextualize subsequent slope estimates.

## Aim 1: Temporal Trends by Age Group

@@

### 1.1 Linear Model - Overall Mortality over time

@@

# Extract model fit statistics
model_fit_lm <- broom::glance(lm_model_year) %>%
  dplyr::mutate(model = "LM: National Year Trend")

# Save as table
saveRDS(model_fit_lm, here::here("results", "tables", "model_fit_national.rds"))

@@

This model provides a first look at overall temporal trends in mortality rates.

The linear regression model estimated a modest but statistically significant annual increase in national hypertension-related CVD mortality rates between 2000 and 2019 (slope = 1.40, p < 2.2e-16). However, as shown in the plot, the fitted trend line is visually flat due to the scale and large variation in county-level rates. The intercept value is not directly interpretable, but reflects the expected mortality rate at year zero in the model's coding. Despite the statistically significant trend, the magnitude of change is small relative to the observed range, suggesting that national mortality rates remained relatively stable over time. This model does not account for age or other demographic factors, motivating the more detailed modeling in the next section.

### 1.2 GLM: Race and Sex as predictors GLM with race and sex as additive predictors to assess overall differences in mortality rates

```{r}
# Examining multiple predictors
glm_model <- glm(mortalities ~ Year + race_ethnicity + sex, data = county_race_sex, family = gaussian())
glm_results <- broom::tidy(glm_model)

@@

The GLM including race/ethnicity and sex as predictors highlights large disparities in predicted mortality rates. Black populations had the highest predicted rates, significantly higher than all other groups (e.g., β = 205.8, p < 2e-16). Hispanic and Asian/Pacific Islander groups had lower mortality, though the Asian effect was not significant (p = 0.35). Males also showed significantly higher mortality than females (β = 20.3, p < 2e-16). Notably, the “Overall” category behaves inconsistently in plots due to it being modeled as a separate group rather than a true average, likely affected by unmodeled age variation. This supports the need to model interactions and stratify further by age.

### 1.3 GLM: Year by Race interaction model: Fit GLM including interaction between year and race to examine race-specific trends over time

@@

saveRDS(county_race_sex, file = here::here("results", "output", "county_race_sex_with_preds.rds"))
```

Including an interaction between year and race revealed significantly divergent temporal trends. Mortality rates among Black individuals showed the steepest annual decline (β = –7.06, p < 0.001), while American Indian and Alaska Native (AI/AN) populations exhibited the largest annual increase (β for year = 3.38, p < 0.001, with positive main effect and significant negative interactions for other groups). Asian and Pacific Islander, Hispanic, and White populations also showed declining trends, though to a lesser extent. These results highlight substantial heterogeneity in how hypertension-related CVD mortality changed over time across racial/ethnic groups.

Given known age-related disparities in cardiovascular risk, we next modeled mortality trends across age groups. This model assesses whether older adults experienced steeper or flatter trajectories over time.

### 1.4: GLM: Year by Age group

GLM with interaction between year and age group to examine differential temporal trends

```{r}
# GLM with interaction: does time trend differ by age group?
glm_age_group <- glm(mortalities ~ Year * age_group, 
                     data = processed_data_no_percent,
                     family = gaussian())

@@

# Extract model fit statistics
model_fit_agegroup <- broom::glance(glm_age_group) %>%
  dplyr::mutate(model = "GLM: Age Group")

# Save as table
saveRDS(model_fit_agegroup, here::here("results", "tables", "model_fit_agegroup.rds"))

@@

This model identifies age-specific trends.Mortality among individuals aged 65+ remains substantially higher and shows modest increases, while younger age groups have lower, flatter trends.
Including an interaction between year and age group revealed that mortality rates are substantially higher in the 65+ age group across all years. While both age groups show increasing mortality over time, the slope is significantly steeper for older adults (interaction term β = 0.224, p < 0.001), suggesting accelerated increases in mortality among those aged 65 and older.

### 1.5: Slope comparison by Age group

@@

# Format slope estimates for reporting
slope_table_age_fmt <- as.data.frame(summary(em_trends_age)) %>%
  dplyr::mutate(
    Estimate = round(Year.trend, 3),
    CI = paste0(round(lower.CL, 2), "–", round(upper.CL, 2)),
    SE = round(SE, 3)) %>%
  dplyr::select(`Age Group` = age_group, Estimate, SE, CI)

# Save as manuscript-ready table
saveRDS(slope_table_age_fmt, here::here("results", "tables", "slope_table_agegroup_fmt.rds"))

@@

# Format pairwise slope differences for age group
slope_contrast_age_fmt <- as.data.frame(summary(slope_contrast, infer = c(TRUE, TRUE))) %>%
  dplyr::mutate(
    Estimate = round(estimate, 3),
    CI = paste0(round(lower.CL, 2), "–", round(upper.CL, 2)),
    SE = round(SE, 3),
    p_value = signif(p.value, 3)
  ) %>%
  dplyr::select(Contrast = contrast, Estimate, SE, CI, p_value)

# Save formatted contrast table
saveRDS(slope_contrast_age_fmt, here::here("results", "tables", "slope_contrast_agegroup_fmt.rds"))

slope_contrast_age_fmt

@@

Estimated yearly changes in mortality rates were significantly higher for adults aged 65+ (slope = 1.26, 95% CI: 1.18–1.34) compared to those aged 35–64 (slope = 1.04, 95% CI: 0.96–1.12). The difference in slope between age groups was statistically significant (difference = -0.224, p < 0.001), indicating a steeper rise in mortality among older adults.

Overall, the results from Aim 1 highlight that age is a significant modifier of temporal trends in hypertension-related cardiovascular mortality. While the overall trend appeared stable at the national level, subgroup analyses revealed that mortality has been increasing more sharply among older adults, and certain racial/ethnic groups show distinct patterns of change over time. These findings justify further exploration of disparities and spatial variation in subsequent aims.

## Aim 2: Spatial Analysis of Hypertension Mortality Rates

## Evaluating spatial distribution and clustering patterns

@@

### 2.1: Static maps by Race:

Converting county data to spatial object using longitude/latitude. 
Creating static spatial maps of county-level mortality rates by race/ethnicity, visualized using spatial coordinates. These provide an overview of geographic variation across U.S. counties.

```{r}
#Convert dataset to an SF spatial object
county_race_sex_sf <- county_race_sex %>%
  st_as_sf(coords = c("X_long", "Y_lat"), crs = 4326, remove = FALSE)

@@

### 2.2 Combined spatial plot

Combining all spatial race-specific maps into a multi-panel figure.

@@

These plots are best viewed as images outside of R due to size.

Static maps show clear geographic variation in hypertension-related CVD mortality. Black and White populations exhibit widespread mortality across the Southeast and Midwest, while Hispanic and Asian populations show more localized clusters, reflecting regional concentration and data availability. The highest mortality hotspots for American Indian and Alaska Native groups are found in the central U.S., consistent with known health disparities.

While visual maps provide insight into geographic differences in mortality, they do not quantify the extent of spatial clustering. To formally assess spatial autocorrelation, we next apply Global Moran’s I and Local Indicators of Spatial Association (LISA) statistics.

### Evaluating spatial clustering of Mortality Rates.
### 2.3: Global Moran’s I – All Races Combined

@@

Running Global Moran's I test to evaluate overall spatial autocorrelation.

```{r spatial-weights global-morans-i, message=FALSE}
# Define file paths for output objrects
knn_path <- here::here("results", "output", "knn_neighbors.rds") # Larger, reusable object
lw_path <- here::here("results", "output", "spatial_weights_list.rds") # Larger, reusable object
moran_path <- here::here("results", "output", "global_morans_i.rds") # Smaller, result-level output

@@

# Extract key statistics from Moran's I test
moran_table <- tibble::tibble(
  Region = "USA (All Counties)",
  Moran_I = round(moran_result$estimate[["Moran I statistic"]], 4),
  Expected_I = round(moran_result$estimate[["Expectation"]], 4),
  SD = round(moran_result$estimate[["Variance"]], 4),
  p_value = signif(moran_result$p.value, 3))

moran_table

# Save to results/tables
saveRDS(moran_table, here::here("results", "tables", "global_morans_i_summary.rds"))
saveRDS(moran_result, file = moran_path)

@@

The Global Moran’s I statistic (I = 0.0435, p < 0.001) indicates weak but statistically significant positive spatial autocorrelation in hypertension-related CVD mortality at the county level. There is weak-to-moderate positive spatial autocorrelation in mortality rates across U.S. counties — counties with high (or low) mortality tend to be near others with similarly high (or low) values. Although the magnitude is small (I = 0.0435), the pattern is considered statistically significant based on the p-value. This suggests that counties with high mortality rates tend to cluster geographically, rather than being randomly distributed. The presence of spatial autocorrelation supports further analysis using Local Moran’s I to identify specific high- and low-mortality clusters.

After confirming statistically significant spatial autocorrelation at the national level, we use Local Moran’s I to identify potential clusters of high or low mortality at the county level.

@@

Running Local Moran's I (LISA) to detect local spatial clusters.

```{r lisa-analysis, message=FALSE}

@@

Creating map of statistically significant LISA clusters (p \< 0.05).

@@

Classifying counties based on LISA cluster type using z-scores.

@@

# Define cluster color palette
cluster_colors <- c(
  "High-High" = "red3",
  "Low-Low" = "blue3",
  "High-Low (Outlier)" = "orange",
  "Low-High (Outlier)" = "purple",
  "Not Significant" = "gray85")

# Plot cluster types
lisa_cluster_plot <- ggplot(county_lisa_coords) +
  geom_point(aes(x = long, y = lat, color = lisa_cluster_type), alpha = 0.7, size = 0.75) +
  scale_color_manual(values = cluster_colors) +
  labs(
    title = "Local Spatial Clusters of Mortality (LISA)",
    subtitle = "High–High = Hotspot, Low–Low = Coldspot, Others = Outliers",
    color = "Cluster Type") +
  coord_equal() +
  theme_minimal(base_size = 12)

@@

ggsave(
  here::here("products", "manuscript", "supplement", "lisa_cluster_type_map.png"),
  plot = lisa_cluster_plot,
  width = 10, height = 6)

Local Moran’s I was computed on the full national dataset using 5-nearest-neighbor spatial weights. Each county was represented by a centroid, and clustering patterns reflect spatial autocorrelation of mortality rates across all years and subgroups.

#### Focusing at a State level Due to the size of the dataset it is challenging to work with the full dataset. 
For convenience and because the SouthEast is historically known to have a high prevalence of cardiovasccular mortalities we will focus on Georgia for this next step. Our previous analysis support that there is a general distribution of cases here, even if there are not distinct hotspots.

Analyzing spatial clusters in Georgia using LISA.

```{r}
# Georgia: Create or load spatial weights
ga_knn_path <- here::here("results", "output", "ga_knn_neighbors.rds")
ga_lw_path  <- here::here("results", "output", "ga_spatial_weights.rds")

@@

if (file.exists(ga_knn_path) && file.exists(ga_lw_path)) {
  message("Reusing saved GA spatial weights and neighbors.")
  ga_knn_nb <- readRDS(ga_knn_path)
  ga_lw <- readRDS(ga_lw_path)
} else {
  message("Generating GA spatial weights.")
  ga_coords <- st_coordinates(county_ga_sf)
  ga_knn_nb <- knn2nb(knearneigh(ga_coords, k = 5))
  ga_lw     <- nb2listw(ga_knn_nb, style = "W")
  saveRDS(ga_knn_nb, file = ga_knn_path)
  saveRDS(ga_lw, file = ga_lw_path)}

@@

cluster_colors <- c(
  "High-High" = "red3",
  "Low-Low" = "blue3",
  "High-Low (Outlier)" = "orange",
  "Low-High (Outlier)" = "purple",
  "Not Significant" = "gray80")

ga_lisa_plot <- ggplot(ga_coords_df) +
  geom_point(aes(x = long, y = lat, color = cluster_type), alpha = 0.8, size = 2) +
  scale_color_manual(values = cluster_colors) +
  labs(
    title = "LISA Cluster Types: Georgia",
    subtitle = "High–High = Hotspot; Low–Low = Coldspot; Others = Outliers",
    color = "Cluster Type") +
  coord_equal() +
  theme_minimal(base_size = 13)

@@

  # Save for reuse
  saveRDS(ms_knn_nb, file = ms_knn_path)
  saveRDS(ms_lw, file = ms_lw_path)}

@@

county_ms_sf <- county_ms_sf %>%
  mutate(
    local_I = ms_local_moran[, 1],
    local_p = ms_p,
    mortality_z = ms_z,
    lag_z = ms_lag_z,
    cluster_type = case_when(
      local_p >= 0.05 ~ "Not Significant",
      mortality_z > 0 & lag_z > 0 ~ "High-High",
      mortality_z < 0 & lag_z < 0 ~ "Low-Low",
      mortality_z > 0 & lag_z < 0 ~ "High-Low (Outlier)",
      mortality_z < 0 & lag_z > 0 ~ "Low-High (Outlier)",
      TRUE ~ NA_character_))

@@

cluster_colors <- c(
  "High-High" = "red3",
  "Low-Low" = "blue3",
  "High-Low (Outlier)" = "orange",
  "Low-High (Outlier)" = "purple",
  "Not Significant" = "gray80")

ms_lisa_plot <- ggplot(ms_coords_df) +
  geom_point(aes(x = long, y = lat, color = cluster_type), alpha = 0.8, size = 2.5) +
  scale_color_manual(values = cluster_colors) +
  labs(
    title = "LISA Cluster Types: Mississippi",
    subtitle = "High–High = Hotspot; Low–Low = Coldspot; Others = Outliers",
    color = "Cluster Type") +
  coord_equal() +
  theme_minimal(base_size = 13)

@@

ggsave(filename = here::here("products", "manuscript", "supplement", "ms_lisa_cluster_plot.png"),
  plot = ms_lisa_plot,
  width = 8,
  height = 6)

@@

# Add LISA values and classification
county_tx_sf <- county_tx_sf %>%
  mutate(
    local_I = tx_local_moran[, 1],
    local_p = tx_p,
    mortality_z = tx_z,
    lag_z = tx_lag_z,
    cluster_type = case_when(
      local_p >= 0.05 ~ "Not Significant",
      mortality_z > 0 & lag_z > 0 ~ "High-High",
      mortality_z < 0 & lag_z < 0 ~ "Low-Low",
      mortality_z > 0 & lag_z < 0 ~ "High-Low (Outlier)",
      mortality_z < 0 & lag_z > 0 ~ "Low-High (Outlier)",
      TRUE ~ NA_character_))

@@

# Cluster colors
cluster_colors <- c(
  "High-High" = "red3",
  "Low-Low" = "blue3",
  "High-Low (Outlier)" = "orange",
  "Low-High (Outlier)" = "purple",
  "Not Significant" = "gray80")

# Plot
tx_lisa_plot <- ggplot(tx_coords_df) +
  geom_point(aes(x = long, y = lat, color = cluster_type), alpha = 0.8, size = 2) +
  scale_color_manual(values = cluster_colors) +
  labs(
    title = "LISA Cluster Types: Texas",
    subtitle = "High–High = Hotspot; Low–Low = Coldspot; Others = Outliers",
    color = "Cluster Type") +
  coord_equal() +
  theme_minimal(base_size = 13)

@@

# Save plot
ggsave(
  filename = here::here("products", "manuscript", "supplement", "tx_lisa_cluster_plot.png"),
  plot = tx_lisa_plot,
  width = 9, height = 6)
```

@@

# Add results to object
county_ky_sf <- county_ky_sf %>%
  mutate(
    local_I = local_moran_ky[, 1],
    local_p = local_moran_ky[, 5],
    mortality_z = scale(mortalities)[, 1],
    lag_z = lag.listw(ky_lw, mortality_z),
    lisa_cluster = case_when(
      local_p > 0.05 ~ "Not Significant",
      mortality_z > 0 & lag_z > 0 ~ "High-High",
      mortality_z < 0 & lag_z < 0 ~ "Low-Low",
      mortality_z > 0 & lag_z < 0 ~ "High-Low (Outlier)",
      mortality_z < 0 & lag_z > 0 ~ "Low-High (Outlier)",
      TRUE ~ NA_character_))

@@

the LISA plot shows spatial patterns across all available data, not just one year

The Local Moran’s I analysis detected significant spatial clustering of mortality rates across U.S. counties. Nationally, over 40% of counties showed statistically significant clustering at p < 0.05. However, most of these represented spatial outliers—counties with high (or low) mortality rates surrounded by contrasting neighbors—rather than concentrated hotspots. Only a small number of “High–High” clusters (hotspots) were identified, indicating that extreme spatial concentrations of high mortality are rare.

Cluster types varied across states:

In Georgia, clustering was driven primarily by High–Low outliers and a few Low–Low coldspots, especially in the northern region.

Mississippi showed a similar pattern, with isolated High–Low counties in the central and southern parts of the state.

In Texas, counties in the western and southern regions showed a mix of Low–Low coldspots and High–Low outliers, with no evidence of strong regional hotspots.

Kentucky had a slightly denser distribution of significant clusters, including more scattered High–Low and Low–Low counties. A few High–High hotspots were detected but did not form a continuous cluster.

Local spatial analysis (LISA) identified a small number of statistically significant clusters in Kentucky, including high-mortality counties adjacent to other high-mortality areas. These patterns support the presence of localized mortality hotspots.

Across all states, the most consistent pattern was the presence of outlier counties—areas with mortality rates that diverged significantly from their neighbors.

@@

### 2.5 Global Moran's I Summary Table

```{r}
# Already saved or placed earlier
extract_moran_values <- function(obj, geo) {
  tibble(
    Geography = geo,
    `Moran's I` = round(obj$estimate[["Moran I statistic"]], 4),
    Expectation = round(obj$estimate[["Expectation"]], 4),
    Variance = round(obj$estimate[["Variance"]], 6),
    `Z Statistic` = round(obj$statistic, 4),
    `p-value` = signif(obj$p.value, 4))}

@@

# Format with gt
moran_gt_tbl <- moran_summary %>%
  gt() %>%
  tab_header(
    title = "Global Moran's I Statistics by Geography",
    subtitle = "Spatial Autocorrelation of Mortality Rates"
  ) %>%
  fmt_number(columns = 2:5, decimals = 4) %>%  # Keep decimals for other metrics
  fmt_scientific(columns = vars(`p-value`), decimals = 2) %>%  # Use sci notation for p-values
  cols_label(
    `Moran's I` = "Moran's I",
    `Z Statistic` = "Z Statistic",
    `p-value` = "p-value"
  ) %>%
  tab_options(
    table.font.size = px(12),
    heading.title.font.size = px(14),
    heading.subtitle.font.size = px(12),
    column_labels.font.weight = "bold")

@@

A summary table of Global Moran’s I statistics for the full U.S. dataset and selected states (GA, MS, TX, KY) is shown below, indicating statistically significant spatial autocorrelation in all geographies examined (p \< 0.001).

Global Moran’s I statistics were statistically significant across all regions analyzed. The USA showed weak but significant positive spatial autocorrelation (I = 0.0435, p < 0.001). Georgia, Mississippi, Texas, and Kentucky also had low Moran’s I values (0.0168–0.0297), but all were statistically significant (p < 0.001), confirming that mortality rates were not randomly distributed in space. These findings support further investigation of local spatial clustering using LISA.

## Aim 3: Demographic Disparities

@@

# Extract model fit metrics
model_fit_race_sex <- broom::glance(glm_model_interaction) %>%
  dplyr::mutate(model = "GLM: Race × Sex × Year")

# Save for results
saveRDS(model_fit_race_sex, here::here("results", "tables", "model_fit_race_sex.rds"))

# Create a table of slopes for each race-sex group 
# Format slope estimates for manuscript table
slope_table_race_sex_fmt <- as.data.frame(summary(em_trends_race_sex)) %>%
  dplyr::mutate(
    Estimate = round(Year.trend, 3),
    CI = paste0(round(lower.CL, 2), "–", round(upper.CL, 2)),
    SE = round(SE, 3)) %>%
  dplyr::select(Race = race_ethnicity, Sex = sex, Estimate, SE, CI)

# Save nicely formatted table
saveRDS(slope_table_race_sex_fmt, here::here("results", "tables", "slope_table_race_sex_fmt.rds"))

#Display table:
#Note: p.value of 0.00e+00 is actually p < 2.2e-16
view(slope_table_race_sex_fmt)

@@

# Create a table comparing slopewise pairwise comparison between groups for each race-sex group
# Format race/sex slope difference table
slope_contrast_race_sex_fmt <- as.data.frame(summary(slope_contrast_race_sex, infer = c(TRUE, TRUE))) %>%
  dplyr::mutate(
    Estimate = round(estimate, 3),
    CI = paste0(round(lower.CL, 2), "–", round(upper.CL, 2)),
    SE = round(SE, 3),
    p_value = signif(p.value, 3)) %>%
  dplyr::select(Contrast = contrast, Estimate, SE, CI, p_value)

#Sort table results by significance (p-value)
slope_contrast_race_sex_fmt_sorted <- slope_contrast_race_sex_fmt %>%
  arrange(p_value, desc(abs(Estimate)))

# Save to results folder
saveRDS(slope_contrast_race_sex_fmt, here::here("results", "output", "slope_contrast_race_sex_fmt.rds"))
saveRDS(slope_contrast_race_sex_fmt_sorted,here::here("results", "output", "slope_contrast_race_sex_sorted.rds"))

# view results from the sorted table 
slope_contrast_race_sex_fmt_sorted

@@

-   Black adults, particularly Black females, showed the steepest and most consistently negative slopes (−3.69 deaths per 100,000 per year, 95% CI: −4.07 to −3.30), indicating large annual declines in mortality. These trends were significantly different from nearly all other groups, with pairwise differences of over 7 points when compared to American Indian and Alaska Native (AIAN) individuals, and around 5.7 points compared to White adults (p \< 2.2e-16).
-   In contrast, American Indian and Alaska Native adults had the most positive slopes (\~+3.38 per year), suggesting a consistent increase in hypertension-related CVD mortality over time. These trends were significantly higher than all other groups, especially when compared to Black, Hispanic, and Asian and Pacific Islander subgroups.
-   White adults had moderately positive slope estimates (\~+1.98 per year), significantly higher than those for Asian, Black, and Hispanic adults. -Hispanic adults and Asian and Pacific Islander adults showed relatively flat to slightly negative trends. These groups had annual change estimates that were not significantly different from zero, with confidence intervals crossing or approaching zero. Differences from the “Overall” group estimates were also generally not statistically significant.
-   Several race/sex combinations (e.g., Black Female vs. White Male; AIAN Male vs. Hispanic Male) yielded statistically significant contrasts, with p \< 2e-16 for many pairwise comparisons, further reinforcing the existence of sharp disparities in trend slopes. These findings suggest that while national mortality may be decreasing on average, substantial heterogeneity by race and sex persists. The increasing trends in mortality among AIAN individuals are particularly concerning, in contrast to strong improvements among Black adults.

While the above results illustrate broad disparities across race and sex, they do not account for variation by age. We next fit an age-adjusted model to stratify mortality trends by age group and explore how disparities unfold differently for younger and older adults.

### 3.2 Demogrpahic disparities by Age group

@@

Overview of the Summary Statistics before modeling.
